A general and efficient representation of Ancestral Recombination Graphs

### 2023-08-29 paper.tex

\section*{Converting an eARG to a gARG}
The process of converting an eARG to a gARG is very similar to
simplifying a prospective gARG with respect to a set of samples,
and highlights some important differences between the encodings.
The process involves two steps, and is illustrated in
Fig~\ref{fig-ancestry-resolution}.

@@

The first step is straightforward
(Figure~\ref{fig-ancestry-resolution}B).
We duplicate the nodes and
edges from the input eARG, and the add inheritance annotations
to the output gARG's edges. If the node is a common ancestor node,
we annotate the single outbound edge with the interval $(0,L]$.
If the node is a recombination node with a breakpoint $x$,
we annotate the outbound edges with the intervals $(0, x]$
and $(x, L]$.
These inheritance annotations are clearly in one-to-one
correspondance with the information in the input eARG,
but they are a superset of the
``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
of the sample.

@@

Fig~\ref{fig-ancestry-resolution}C shows the graph that we obtain
by resolving the ancestral material transmitted along each edge
(using the ``simplify'' algorithm), and highlights some important
differences between the eARG and gARG encodings.
Firstly, we can see that some nodes and edges are removed entirely
from the graph. The ``grand MRCA'' $q$ is ommitted from the
fully resolved gARG because all segments of the genome have
fully coalesced before it is reached. Likewise, the edge
between $g$ and $j$ is omitted because the recombination
event at position $5$ fell in non-ancestral material.
More generally, we can see that the fully resolved
gARG encoding of Fig~\ref{fig-ancestry-resolution}C
allows for ``local'' inspection
of an ARG in a way that is not possible by storing
common ancestor and recombination events in an eARG. Because
the ancestral material is stored with each edge, the effects
of multiple events can be reasoned about at once. Many algorithms
that we wish to perform on an ARG will require resolving
the genetic material with respect to a sample.
% this is weak, but would be good to emphasise this point.
% you just end up running simplify lots of times, in your
% downstream algs
The gARG encoding
allows us to do this once and to store the result, whereas
the eARG encoding requires us to repeat the process on
each traversal.

@@

Fig.~\ref{fig-simplification} show a series of successive steps of
simplification, starting with a complete gARG simulated under a
backwards-in-time Wright-Fisher
model, shown both as a graph on the left, and the corresponding sequence of 6
local trees on the right.
For example, consider the path above node
\textsf{a} in the first tree (covering positions 30 to 40)
in Fig.~\ref{fig-simplification}A.
Before \textsf{a} meets its common ancestor with the other samples at
node \textsf{q}, it passes through \textsf{e}, \textsf{j}, \textsf{m}, \textsf{n} and
\textsf{p}, which in this region of the genome are all unary nodes.
This reflects the passage of this span of genome through the ARG towards the root.
Such unary nodes in a local tree are not usually considered important;
% because they provide little information and are not
% detectable by information from any given individual tree.
nodes usually mark the points where samples coalesce in a
 \emph{common} ancestor. For most purposes, we would therefore consider a path directly
joining sample \textsf{a} to node \textsf{q} to be equivalent to the path
containing the intermediate unary nodes.

\begin{figure}
\centering
\vspace{5em}
\includegraphics[width=\linewidth]{illustrations/simplification}
\caption{\label{fig-simplification}
Levels of ARG simplification.
In all cases the graph visualization is shown on the left
and the corresponding local trees on the right.
Nodes are coloured by the number of parents and shaded
according to the proportion of their span over which they are coalescent.
(A) A gARG simulated from a diploid Wright-Fisher
model. Note that the genome \textsf{k} has three children, and the genomes
\textsf{e} and \textsf{g} simultaneously have multiple parents and multiple
children.
(B) Simplified to remove all
% trying out this terminology - can define in the text
1-connected graph components (e.g., diamonds such as \textsf{jlnm}).
(C) Remove nodes that are unary in all local trees. This removes both recombinant nodes
that have only one child (e.g. \textsf{n}) and nodes that have more
than one child (i.e. are ``common ancestors'') but never represent coalescences
in the local trees (e.g. \textsf{r}).
(D) Rewrite edges to bypass any unary nodes in local trees. Note that this loses lineage
information such as the fact that \textsf{k} is connected to \textsf{a} via \textsf{e}.
}
\end{figure}

The first level of simplification that we can perform is to remove
graph topology that is invisible to the samples.
An example of such topology is a
``diamond''~\citep{rasmussen2014genome}
in which the two parent nodes of a recombination immediately
join again into a common ancestor (e.g.~\textsf{j}, \textsf{l}, \textsf{m}
and \textsf{n} in Fig.~\ref{fig-simplification}A).
Unless we are specifically
interested in the recombination event or these ancestral genomes,
there is no information in this topology and the diamond can be
replaced by a single edge. More generally, any
subgraph that is singly-connected in both the leafward and
rootward direction (a ``super-diamond'') is non-identifiable and can be
replaced by one edge. This definition includes the case
of a node that has one inbound and one outbound edge, such as
nodes \textsf{f} and \textsf{h}.
Fig.~\ref{fig-simplification}B shows the result of this type of
graph topology simplification.

Simplifying away diamonds will remove many unary nodes from the
local trees, but there can still be nodes that are unary in all
of the local trees. In particular, a node can represent a recombinant
with multiple parents in the graph but only a single child (e.g.\ node \textsf{n}
in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
multiple children in the graph but in which no coalescence takes place
in the local trees
(node \textsf{r} in Fig.~\ref{fig-simplification}B).
% point out that this is why Hudson refers to CA nodes, not coalescence nodes
Such nodes are not singly connected in the graph, but are nevertheless unary in
all of the local trees.
The operation to remove them
therefore requires knowledge not just of the graph topology but also of the
ancestral material associated with the edges.
As we see in Fig.~\ref{fig-simplification}C,
removal of recombinant nodes can produce graph nodes with
more than two parents (e.g.~node \textsf{e}); and likewise, removal of
common ancestor but non-coalescent nodes can produce graph nodes with
more than two children (e.g.~node \textsf{s}). These represent a ``stacking up'' of
multiple---often indistinguishable---events into a single node.
% something which is impossible to encode in an eARG.

% JK: Text moved from an earlier section. See if we want to reuse this here
% if not already said.
% Indeed a single node can have both multiple children
% and multiple parents. This would be the case in Fig~\ref{fig-arg-in-pedigree} if,
% for instance, node \textsf{a} were to have two children.
% % It is a shame that this is not actually shown in fig 3, but I don't think
% % we can do this unless we have 5 generations
% Moreover, a single node can have more than 2 children, and (as we shall see in
% the ``\nameref{ARG_simplification}'' section), more than 2 parents. Another way to
% state this is that transitions between nodes in a gARG can encapsulate multiple
% events in an eARG.

The remaining nodes are MRCAs of some subset of the samples
at \emph{some} positions along the genome. We still have
some unary nodes in the local trees, but these nodes will
correspond to a coalescence in at least one other
local tree. For example, node  \textsf{k} is unary in the second tree
of Fig.~\ref{fig-simplification}C, but is either binary
or ternary in all subsequent trees (recall this is a Wright-Fisher
simulation). The final level of simplification is to alter the edge annotations
such that, although no nodes are removed from the graph, all
unary nodes disappear from the local trees (Fig.~\ref{fig-simplification}D).
Note that although this last stage produces simpler local trees, by
removing information about the exact paths taken by lineages through
the graph, we lose potentially useful information about shared edges
between trees. The extent to which this information can be inferred
and utilised is an interesting open question.

@@

The ARGs in Fig.~\ref{fig-simplification} contain different
numbers of local trees ($6$, $5$, $5$ and $4$ for A through
D, respectively). When we move from A to B the local trees
for the intervals $(44,61]$ and $(61,87]$ are merged because
the only differences between them are their journeys through
nodes \noderef{l} and \noderef{m}. These nodes that participated
in the diamond are removed from the ARG, and we have lost
all information about the corresponding recombination at
position 61. Other nodes (e.g.\ \noderef{o} and \noderef{p})
have also been removed but these represent the \emph{parents}
of recombinants. The recombinant nodes themselves
(e.g. \noderef{n}) are still present, and represent precise
information about the time, genomic location and tree
branches involved
in the recombination event.

@@

The point of these examples is to emphasise that
information about recombination in an ARG is not ``all
or nothing'': there is a continuuum of detail
about recombination, in terms of genomic location, timing
the ancestral lineages involved, that we can describe.
While there has been a recent tendancy to emphasise the
importance of ``full ARGs''
\citep[e.g.][]{deng2021distribution,brandt2021evaluation,
rasmussen2022espalier}
which contain complete information about all recombination
events, it is reasonable to question the value of
such precise information and whether it is knowable.
[SOME MORE STUFF]

>>>>

\section*{Converting an eARG to a gARG}
Any eARG can be converted to a gARG without loss of information,
and the conversion process is very similar to
simplifying a prospective gARG with respect to a set of samples.
The process involves two steps, and is illustrated in
Fig~\ref{fig-ancestry-resolution}.
The first step is straightforward
(Figure~\ref{fig-ancestry-resolution}B).
We duplicate the nodes and
edges from the input eARG, and the add inheritance annotations
to the output gARG's edges. If the node is a common ancestor node,
we annotate the single outbound edge with the interval $(0,L]$,
for a genome of length $L$.
If the node is a recombination event with a breakpoint $x$,
we annotate the outbound edges with the intervals $(0, x]$
and $(x, L]$.
These inheritance annotations are clearly in one-to-one
correspondance with the information in the input eARG,
but they are a superset of the
``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
of the sample.

@@

@@

Fig~\ref{fig-ancestry-resolution}C shows the graph that we obtain
by resolving the ancestral material transmitted along each edge
(using the ``simplify'' algorithm), and highlights some important
differences between the eARG and gARG encodings.
Firstly, we can see that some nodes and edges are removed entirely
from the graph. The ``grand MRCA'' \noderef{q} is omitted from the
fully resolved gARG because all segments of the genome have
fully coalesced before it is reached. Likewise, the edge
between \noderef{g} and \noderef{j} is omitted because the recombination
event at position $5$ (represented by node \noderef{g})
fell in non-ancestral material.
More generally, we can see that the fully resolved
gARG encoding of Fig~\ref{fig-ancestry-resolution}C
allows for ``local'' inspection
of an ARG in a way that is not possible by storing
common ancestor and recombination events in an eARG. Because
the ancestral material is stored with each edge, the
cumulative effects of events over time can be reasoned
about at once. Many computations
that we wish to perform on an ARG will require resolving
the genetic material with respect to a sample.
% this is weak, but would be good to emphasise this point.
% you just end up running simplify lots of times, in your
% downstream algs
The gARG encoding
allows us to do this once and to store the result, whereas
the eARG encoding requires us to repeat the process on
each traversal.

@@

\begin{figure}
\centering
\vspace{5em}
\includegraphics[width=\linewidth]{illustrations/simplification}
\caption{\label{fig-simplification}
Levels of ARG simplification.
In all cases the graph visualization is shown on the left
and the corresponding local trees on the right.
Nodes are coloured by the number of parents and shaded
according to the proportion of their span over which they are coalescent.
(A) A gARG simulated from a diploid Wright-Fisher
model. Note that the genome \noderef{k} has three children, and the genomes
\noderef{e} and \noderef{g} simultaneously have multiple parents and multiple
children.
(B) Simplified to remove all
singly-connected graph components (e.g., diamonds such as \noderef{jlnm}).
(C) Remove nodes that are unary in all local trees. This removes both recombinant nodes
that have only one child (e.g.\ \noderef{n}) and nodes that have more
than one child (i.e. are ``common ancestors'') but never represent coalescences
in the local trees (e.g.\ \noderef{r}).
(D) Rewrite edges to bypass any unary nodes in local trees. Note that this loses lineage
information such as the fact that \noderef{k} is connected to \noderef{a} via \noderef{e}.
}
\end{figure}

Fig.~\ref{fig-simplification} show a series of successive steps of
simplification, starting with a complete [TODO be more concrete about
what is stored here; ``complete'' isn't very helpful] gARG simulated under a
backwards-in-time Wright-Fisher
model, shown both as a graph on the left, and the corresponding sequence of
local trees on the right.
In Fig.~\ref{fig-simplification}A the local trees (right)
contain many unary nodes, and as we simplify more and more
the local trees contain fewer
unary nodes (Fig.~\ref{fig-simplification}B,C) until
we reach Fig.~\ref{fig-simplification}D in which the local trees
have no unary nodes.
Unary nodes in a local tree arise from
genetic inheritance
passing through an ARG node in which no coalescence
occurs in the corresponding genome interval.
The distinction between the ``common ancestry'' of two or more genomes
in an ancestral genome and the ``coalescence'' which may or may
not occur in the local trees is
important~\citep{hudson1983testing,kelleher2016efficient}.
Consider \noderef{e} in in Fig.~\ref{fig-simplification}A,
for example. We can see from the graph that it is a common
ancestor of  samples \noderef{a} and \noderef{b}, but
it does not correspond to any coalescence in the
local trees to the left of position $44$, and is a unary
node in these three trees. Unary nodes in the local trees
arise from a variety of processes, and would not usually
be considered significant. We are usually interested in tree
nodes that have at least two children, because these
are in principle detectable from mutational differences
arising in the respective subtrees.

% \noderef{a} in the leftmost tree (covering positions 0 to 30)
% Consider, for example, the path ancestral to node
% \noderef{a} in the leftmost tree (covering positions 0 to 30)
% in Fig.~\ref{fig-simplification}A.
% We can see from the graph that
% % Before \noderef{a} meets its common ancestor with the other samples at
% % node \noderef{q}, it passes through \noderef{e}, \noderef{j}, \noderef{m}, \noderef{n} and
% % \noderef{p}, which in this region of the genome are all unary nodes.
% % Node \noderef{e}

% Such unary nodes in a local tree are not usually considered important;
% % because they provide little information and are not
% % detectable by information from any given individual tree.
% nodes usually mark the points where samples coalesce in a
%  \emph{common} ancestor. For most purposes, we would therefore consider a path directly
% joining sample \noderef{a} to node \noderef{q} to be equivalent to the path
% containing the intermediate unary nodes.

The first level of simplification that we can perform is to remove
graph topology that is invisible to the samples.
An example of such topology is a
``diamond''~\citep{rasmussen2014genome}
in which the two parent nodes of a recombination immediately
join again into a common ancestor (e.g.~\noderef{j}, \noderef{l}, \noderef{m}
and \noderef{n} in Fig.~\ref{fig-simplification}A).
Unless we are specifically
interested in the recombination event or these ancestral genomes,
there is no information in this topology and the diamond can be
replaced by a single edge. More generally, any
subgraph that is singly-connected in both the leafward and
rootward direction (a ``super-diamond'') is non-identifiable and can be
replaced by one edge. This definition includes the case
of a node that has one inbound and one outbound edge, such as
nodes \noderef{f} and \noderef{h}.
Fig.~\ref{fig-simplification}B shows the result of this type of
graph topology simplification.

Simplifying away diamonds will remove many unary nodes from the
local trees, but there can still be nodes that are unary in all
of the local trees. In particular, a node can represent a recombinant
with multiple parents in the graph but only a single child (e.g.\ node \noderef{n}
in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
multiple children in the graph but in which no coalescence takes place
in the local trees
(node \noderef{r} in Fig.~\ref{fig-simplification}B).
% point out that this is why Hudson refers to CA nodes, not coalescence nodes
Such nodes are not singly connected in the graph, but are nevertheless unary in
all of the local trees.
The operation to remove them
therefore requires knowledge not just of the graph topology but also of the
ancestral material associated with the edges.
As we see in Fig.~\ref{fig-simplification}C,
removal of recombinant nodes can produce graph nodes with
more than two parents (e.g.~node \noderef{e}); and likewise, removal of
common ancestor but non-coalescent nodes can produce graph nodes with
more than two children (e.g.~node \noderef{s}). These represent a ``stacking up'' of
multiple---often indistinguishable---events into a single node.
% something which is impossible to encode in an eARG.

% JK: Text moved from an earlier section. See if we want to reuse this here
% if not already said.
% Indeed a single node can have both multiple children
% and multiple parents. This would be the case in Fig~\ref{fig-arg-in-pedigree} if,
% for instance, node \noderef{a} were to have two children.
% % It is a shame that this is not actually shown in fig 3, but I don't think
% % we can do this unless we have 5 generations
% Moreover, a single node can have more than 2 children, and (as we shall see in
% the ``\nameref{ARG_simplification}'' section), more than 2 parents. Another way to
% state this is that transitions between nodes in a gARG can encapsulate multiple
% events in an eARG.

The remaining nodes are MRCAs of some subset of the samples
at \emph{some} positions along the genome. We still have
some unary nodes in the local trees, but these nodes will
correspond to a coalescence in at least one other
local tree. For example, node  \noderef{k} is unary in the second tree
of Fig.~\ref{fig-simplification}C, but is either binary
or ternary in all subsequent trees (recall this is a Wright-Fisher
simulation). The final level of simplification is to alter the edge annotations
such that, although no nodes are removed from the graph, all
unary nodes disappear from the local trees (Fig.~\ref{fig-simplification}D).
Note that although this last stage produces simpler local trees, by
removing information about the exact paths taken by lineages through
the graph, we lose potentially useful information about shared edges
between trees. The extent to which this information can be inferred
and utilised is an interesting open question.

@@

The ARGs in Fig.~\ref{fig-simplification} contain different
numbers of local trees ($6$, $5$, $5$ and $4$ for A through
D, respectively). When we move from A to B the local trees
for the intervals $(44,61]$ and $(61,87]$ are merged because
the only differences between them are their paths through
nodes \noderef{l} and \noderef{m}. These nodes that participated
in the diamond are removed from the ARG, and we have lost
all information about the corresponding recombination at
position 61. Other nodes (e.g.\ \noderef{o} and \noderef{p})
have also been removed but these represent the \emph{parents}
of recombinants. The recombinant nodes themselves
(e.g. \noderef{n}) are still present, and represent precise
information about the time, genomic location and tree
branches involved
in the recombination event.

@@

The point of these examples is to emphasise that
information about recombination in an ARG is not ``all
or nothing'': there is a gradation of detail
about recombination, in terms of genomic location, timing,
and the ancestral lineages involved.
While there has been a recent tendancy to emphasise the
importance of ``full ARGs''
\citep[e.g.][]{deng2021distribution,brandt2021evaluation,
rasmussen2022espalier}
which contain complete information about all recombination
events, it is reasonable to question the value of
such precise information and whether it is knowable.
[SOME MORE STUFF]
